raptr: Representative and Adequate Prioritisation Toolkit in R

### 2016-10-12 code/rmarkdown/article.Rmd
+ try using aligned environment
+ attempt 2
+ aligned attempt 2
+ attempt 3
+ assorted improvements after jonathons comments

1. An underlying aim in conservation planning is to maximise the long-term persistence of biodiversity. To fulfill this aim, the processes the sustain biodiversity must be conserved. Many biodiversity processes can be preserved by conserving a representative sample of each biodiversity feature (species or ecosystem). This idea has for decades been encapsulated in the principle of \"representativeness\". For example, representing the range of genetic diversity or ability to persist in a range of climatic conditions might be crucial for the long-term persistence of a species in a world where environmental change is accelerating. Yet current approaches overwhelmingly focus on \"adequacy\", the concept of ensuring a target proportion of the distribution of a biodiversity feature is reserved.

2. To address this issue, we developed the \\texttt{raptr: Representative and Adequate Prioritisation Toolkit in R} to guide reserve selection using explicit targets for representing biodiversity processes as well as area protection targets. Users set \"space targets\" to secure a representative sample of each feature within an attribute space that reflects a biodiversity process (eg. an attribute space expressing variation in genetic characteristics between individuals in an area, or variation in climatic conditions between inhabited areas). We explored the functionality of this package using a simulated study and two case studies. In each study, we generated solutions that secured a proportion of the species' distributions using amount targets--representing conventional reserve selection--and compared them with solutions generated using both amount and space targets.

3. We show that explicitly considering biodiversity processes in reserve selection can substantially change the solution. The simulations suggest that including representativeness in conservation planning is important where biodiversity features have multi-model distributions in an attribute space. The case studies show that explicitly considering space-based targets for biodiversity processes can return solutions that could be far more effective in achieving long term biodiversity persistence and do not require much more area than traditional approaches.

@@

# Introduction
<!-- conservation and reserve networks -->
Perhaps the most fundamental aim of conservation is to maximise the long-term persistence of biodiversity [@r423; @r255]. To achieve this, conservation actions must preserve biodiversity patterns (eg. populations, species, ecosystems), but also crucially the processes that sustain them. One of the major tangible achievements of modern conservation has been the act of setting aside areas for preservation [@r440]. Reserve networks buffer species from gross threatening processes and set the stage for enhanced management interventions [@r465]. However, the resources available for conservation action are limited, and so reserve networks must be sited in places that satisfy conservation objectives for minimal cost [@r255]. To achieve this, reserve selection is often formulated as an optimisation problem and then solved to identify cost-effective candidate reserve systems [termed prioritisations; @r255]. Many decision support tools have been developed to help identify cost-effective solutions [eg. \texttt{ ConsNet}, @r446; \texttt{ Marxan}, @r22; \texttt{ Zonation}, @r429]. Typically, these tools are used to deliver a prioritisation that secures a set of species of interest to the conservation planner (termed target species or features). Given a set of species, decision makers can use these tools to preserve biodiversity patterns by generating a prioritisation that secures an adequate proportion of each species' range. However, to achieve effective conservation, reserve networks must do more than cover enough area.

<!-- biodiversity processes are important -->
Prioritisations must capture the ecological and evolutionary processes that sustain biodiversity [@r3; @r255]. Ecological processes, such as predator-prey interactions, pollination, and decomposition, are required for biodiversity to persist over short time-scales. Typically, they operate over small geographic domains--with exceptions such as migration and refugial habitats--and can be preserved using suitably large planning units [@r446] that each contain a discrete unit of habitat [@r448]. On the other hand, evolutionary processes are required for biodiversity to persist over long time-scales, and they typically operate over large geographic domains. Protected areas must represent adaptive evolutionary processes to foster resilience against environmental change [eg. climate change; @r247]. Protected areas must also preserve neutral evolutionary processes [@r5; @r67; @r63], arising from restriction of gene flow between populations. They are important for maintaining genetic diversity, and avoiding inbreeding depression. In recent decades, a wealth of data relevant to evolutionary processes has become available to conservation planners [eg. bioclimatic, genetic, phylogenetic, and trait data; @r33; @r460; @r453]. This boom has resulted in significant progress towards evolutionary processes into conservation planning at the infra- and inter-specific level [eg. @r68; @r470; @r469]. Yet such data are rarely used to guide conservation planning [@r61]. This is perhaps in large part due to the fact that existing reserve selection tools focus on capturing either biodiversity patterns or processes--but not both.

<!-- parametrisation of biodiversity patterns and processes in conventional conservation planning tools -->
Existing conservation planning tools are generally not well-suited for representing intra-specific biodiversity patterns and processes. To preserve the biodiversity processes that underpin a given species, a prioritisation must capture a representative sample of each species. For instance, to preserve predator-prey interactions, a prioritisation must preserve individuals from each species in the same areas [@r471]. To preserve adaptive evolutionary processes, a prioritisation must preserve the adaptive landscape of each species--populations experiencing different selection pressures [@r63]. To preserve neutral evolutionary processes, a prioritisation must secure individuals descended from each of the genetic lineages that comprise each species [@r5]. Planning tools that focus on such biodiversity processes [eg. \texttt{Diversity}; @r203] are unable to accommodate species-specific targets to ensure that prioritisations secure biodiversity features. Similarly, planning tools that focus on biodiversity patterns (eg. \texttt{Marxan}) cannot easily accomodate biodiversity processes. Previous attempts to accommodate processes into multi-species conservation planning have partitioned species' ranges into different groups, for example at habitat discontinuities [e.g. @r2] as a pre-processing step to render a new set of "pseudo-species". However, this approach assumes that the biodiversity processes that operate across a species' range can be readily split into discrete units. Yet data on biodiversity processes is often continuous, hierarchical, and hyper-dimensional, and so often cannot be reduced to a few categories without significant information loss [@r218].

<!-- aims of this paper -->
One of the key issues in reserve selection is the lack of a unifying decision support tool that can accommodate data on biodiversity patterns and processes in a multi-species context. To begin to fill this gap, we present the \texttt{raptr R} package. This \texttt{R} package uses novel formulations of the reserve selection problem to provide decision makers with the tools to generate prioritisations that preserve both biodiversity patterns and processes. We aim to define the concepts behind the problem formulations. Furthermore, we aim to explore the functionality of the \texttt{raptr R} package by applying one of the formulations to a set of simulated species and two case studies.

@@

The binary genetic data was ordinated into a two-dimensional space that described the main different between individuals (stress = `r round(cs2.spp.nmds.LST[[1]]$stress,2)`). These values were then averaged to the planning unit level to describe the typical genetic characteristics of individuals in each planning unit (Figs. 7c--7d). Planning units located near each were found to contain individuals with similar genetic characteristics (Moran's I: NMDS axis 1, $I = `r round(cs1.nmds.MoransI[[1]][[1]]$observed,2)`$, $P `r pvalString(cs1.nmds.MoransI[[1]][[1]]$p.value)`$; Moran's I: NMDS axis 2, $I = `r round(cs1.nmds.MoransI[[1]][[2]]$observed,2)`$, $P `r pvalString(cs1.nmds.MoransI[[1]][[2]]$p.value)`$). In terms of the average genetic characteristics of individuals in the 	planning units, they tended to cluster into two main groups, with evidence of within-group structure inside the larger group (Fig 8c). This analysis supports previous work by Alvarez _et al._ [-@r461] who also found evidence of genetic structure within this species.

>>>>

1. An underlying aim in conservation planning is to maximise the long-term persistence of biodiversity. To fulfill this aim, the processes that sustain biodiversity must be conserved. Many biodiversity processes can be preserved by conserving a representative sample of each biodiversity feature (species or ecosystem). This idea has for decades been encapsulated in the principle of \"representativeness\". For example, representing the range of genetic diversity or ability to persist in a range of climatic conditions might be crucial for the long-term persistence of a species in a world where environmental change is accelerating. Yet current approaches overwhelmingly focus on \"adequacy\", the concept of ensuring a target proportion of the distribution of a biodiversity feature is reserved.

2. To address this issue, we developed the \\texttt{raptr: Representative and Adequate Prioritisation Toolkit in R} to guide reserve selection using conventional area protection targets as well as explicit targets for representing intra-specific biodiversity processes. Users set \"amount targets\" to preserve an adequate proportion of habitat for each species. They also set \"representation targets\" to secure a representative sample of variation within features reflect a biodiversity process (eg. genetic characteristics between individuals in areas, or variation in climatic conditions between inhabited areas). We explored the functionality of this package using a simulated study and two case studies. In each study, we generated solutions that secured a proportion of the species' distributions using amount targets--representing conventional reserve selection--and compared them with solutions generated using both amount and space targets.

3. We show that explicitly considering biodiversity processes in reserve selection can substantially change the solution. The simulations suggest that including representativeness in conservation planning is important where biodiversity features have multi-modal distributions in an attribute space. The case studies show that explicitly considering representation-based targets for biodiversity processes can return solutions that could be far more effective in achieving long term biodiversity persistence and do not require much more area to be reserved than traditional approaches.

@@

# Introduction
<!-- conservation and reserve selection -->
Perhaps the most fundamental aim of conservation is to maximise the long-term persistence of biodiversity [@r423; @r255]. To achieve this, conservation actions must preserve biodiversity patterns (eg. populations, species, ecosystems), but also crucially the processes that sustain them. One of the major tangible achievements of modern conservation has been the act of setting aside areas for preservation [@r440]. Reserve networks buffer species from gross threatening processes and set the stage for enhanced management interventions [@r465]. Since the resources available for conservation action are limited, protected area networks must be sited in places that satisfy conservation objectives for minimal cost [@r255]. However, despite the importance of biodiversity processes, current approaches for incorporating them into reserve selection are limited.

<!-- biodiversity processes are important -->
Prioritisations must capture the ecological and evolutionary processes that sustain biodiversity [@r3; @r255; @r474]. Ecological processes are important for biodiversity to persist over short time-scales. For instance, predator-prey processes can mediate ecosystem structure and function through top-down control [@rXXX]. Additionally, migratory processes  allow many species to access suitable habitat for breeding [@rXXX]. On the other hand, evolutionary processes are important for biodiversity to persist over long time-scales. Adaptive evolutionary processes foster resilience against environmental change [@r3; @r43], and neutral evolutionary processes maintain genetic diversity [@r5; @r67; @r63]. In recent decades, a wealth of data relevant to biodiversity processes has become available to conservation planners [eg. bioclimatic, distribution, genetic, phylogenetic, and trait data; @rXXX; @r33; @r460; @r453]. This boom has resulted in significant progress towards incorporating both ecological and evolutionary processes into conservation planning [@r473; @r474; @r448; @r2; @r68; @r470; @r469; @r247; @r488; @r468; @r407].

<!-- current methods for preserving biodiversity processes -->
Many decision support tools have been developed to help identify cost-effective solutions [eg. \texttt{ConsNet}, @r446; \texttt{Marxan}, @r22; \texttt{Zonation}, @r429]. Typically, these tools are used to deliver a prioritisation that secures a set of species of interest to the conservation planner (termed target species or features). Given a set of species, decision makers use these tools to generate a prioritisation that secures an adequate proportion of each species' range. To accommodate biodiversity processes, current methods involve partitioning species' ranges into different groups as a pre-processing step to render a new set of "pseudo-species" [eg. at habitat discontinuities; @r2]. However, this approach suffers two main limitations. First, splitting species' ranges into discrete groups is not a trivial task. Data on biodiversity processes are often continuous, hierarchical, and hyper-dimensional, and so may not be able to be reduced to a few groups without significant information loss [@r218]. Second, by splitting the proxy variable into groups, the decision maker cannot directly assess how well a prioritisation actually samples variation in the original proxy variable. Simply setting targets for representing each of the pseudo-species does not necessarily guarantee that the original proxy variable been well sampled--and by extension that the biodiversity process has been preserved.

<!-- aims of this paper -->
One of the key issues in reserve selection is the lack of a unifying decision support tool that can accommodate data on biodiversity patterns and processes. To begin to fill this gap, we present the \texttt{raptr R} package. This \texttt{R} package uses novel formulations of the reserve selection problem to provide decision makers with the tools to generate prioritisations that preserve both biodiversity patterns and processes. Here, we aim to define the concepts that underpin this \texttt{R} package and explore its functionality.

@@

The binary genetic data was ordinated into a two-dimensional space that described the main different between individuals (stress = `r round(cs2.spp.nmds.LST[[1]]$stress,2)`). These values were then averaged to the planning unit level to describe the typical genetic characteristics of individuals in each planning unit (Figs. 7c--7d). Planning units located near each other were found to contain individuals with similar genetic characteristics (Moran's I: NMDS axis 1, $I = `r round(cs1.nmds.MoransI[[1]][[1]]$observed,2)`$, $P `r pvalString(cs1.nmds.MoransI[[1]][[1]]$p.value)`$; Moran's I: NMDS axis 2, $I = `r round(cs1.nmds.MoransI[[1]][[2]]$observed,2)`$, $P `r pvalString(cs1.nmds.MoransI[[1]][[2]]$p.value)`$). In terms of the average genetic characteristics of individuals in the 	planning units, they tended to cluster into two main groups, with evidence of within-group structure inside the larger group (Fig 8c). This analysis supports previous work by Alvarez _et al._ [-@r461] who also found evidence of genetic structure within this species.
